---
title: "MfE Job Interview Task (David Hodge)"
output: pdf_document
---
1. 

```{r, echo=FALSE, message=FALSE, warning=F, fig.height=4}
library(tidyverse)
library(readxl)
library(purrr)
library(ggblanket)

get_ghg <- function(x) {
  read_excel("data/Scenarios-dataset-2021-final-advice.xlsx", x) |>
    select(-c(3, 4)) |>
    fill(1:2) |>
    rename(cat1 = x, cat2 = ...2, cat3 = ...5) |>
    filter(!is.na(cat3)) |>
    group_by(cat1, cat2, cat3) |>
    tidyr::pivot_longer(
      cols = `1990`:`2050`,
      names_to = "year",
      values_to = "ktco2e") |>
    mutate(across(year, \(x) ymd(paste0(x, "0101")))) |>
    mutate(cat0 = x) |>
    select(cat0, everything()) |>
    ungroup()
}

scenarios <- excel_sheets("data/Scenarios-dataset-2021-final-advice.xlsx")[3:10]

ghg <- map(scenarios, get_ghg) |>
  purrr::list_rbind()

#question1
scenarios1 <-
  c("Demonstration path",
    "Headwinds",
    "Tailwinds",
    "Current Policy Reference")

ghg |>
  filter(
    cat0 %in% scenarios1,
    cat1 == "Summary of emissions",
    cat2 == "Summary emissions by GHG Inventory classification (ktCO2e)",
    cat3 == "Net") |>
  filter(between(year, ymd("2020-01-01"), ymd("2050-01-01"))) |>
  mutate(cat0 = factor(cat0, scenarios)) |>
  gg_line(
    x = year,
    y = ktco2e,
    col = cat0,
    y_include = 0,
    y_title = "Net emissions (kt CO\u2082e)",
    col_title = "Scenerio",
    title = "Climate Commission modelled net emissions under different scenarios, 2020\u20132050",
    col_legend_ncol = 2,
    pal = pal_default(5)[c(2, 3, 4, 1)],
    size = 1,
    y_breaks = scales::breaks_width(10000))
```


2. 
```{r, echo=FALSE, message=FALSE, warning=F}
scenarios2 <-
  c("Demonstration path",
    "Current Policy Reference")

emissions_change <- ghg |>
  filter(
    cat0 %in% scenarios2,
    cat1 == "Summary of emissions",
    cat2 == "Summary biogenic CH4 by GHG Inventory classification (ktCO2e)",
    cat3 == "Net") |>
  filter(between(year, ymd("2022-01-01"), ymd("2025-01-01"))) |>
  group_by(cat0) |>
  summarise(across(ktco2e, sum)) |>
  pivot_wider(names_from = cat0,
              values_from = ktco2e,
              names_repair = snakecase::to_snake_case) |>
  mutate(emissions_change = demonstration_path - current_policy_reference) |> 
  pull(emissions_change) |> 
  round(1)
```

Biogenic methane emissions are reduced by `r abs(emissions_change)` ktCO2e over the 1st emissions budget period (i.e. 2022-2025) compared to the Current Policy Reference scenario.

3. 

Current Policy Scenario with project reductions and subsequent emissions in the 1st emissions budget period (2022-2025):

```{r, echo=FALSE, message=FALSE, warning=F}
#question 3
project_cpr <- ghg |>
  filter(
    cat0 %in% "Current Policy Reference",
    cat1 == "Waste",
    cat2 == "Waste emissions by gas (ktCO2e)",
    cat3 == "CH4") |>
  filter(between(year, ymd("2022-01-01"), ymd("2030-01-01")))

#30% emissions reduction in 2030
reduction_2030 <- project_cpr |>
  filter(year == ymd("2024-01-01")) |>
  pull(ktco2e) * -0.33

#emissions reduction in 2024 to allow for doubling each year until 2030 value

# x * 2 ^ 6 = 967
reduction_2024 <- reduction_2030 / 2 ^ 6

#emissions reductions by year between 2022 and 2030
reductions <- c(
  0,
  0,
  reduction_2024,
  reduction_2024 * 2,
  reduction_2024 * 2 ^ 2,
  reduction_2024 * 2 ^ 3,
  reduction_2024 * 2 ^ 4,
  reduction_2024 * 2 ^ 5,
  reduction_2024 * 2 ^ 6)

#project reductions and emissions in the 1st commitment period (2022-2025)
project_cpr |>
  add_column(reductions) |>
  mutate(ktco2e_new = ktco2e + reductions) |>
  filter(between(year, ymd("2022-01-01"), ymd("2025-01-01"))) |>
  summarise(across(c(reductions, ktco2e_new), sum)) |>
  rename(reductions_ar4 = reductions,
         emissions_ar4 = ktco2e_new) |>
  mutate(reductions_ar5 = reductions_ar4 / 25 * 28,
         emissions_ar5 = emissions_ar4 / 25 * 28)
```

Demonstration emissions in the 1st emissions budget period for comparison (2022-2025):

```{r, echo=FALSE, message=FALSE, warning=F}
#compare to Demonstration Pathway equivalent emissions
ghg |>
  filter(
    cat0 %in% "Demonstration path",
    cat1 == "Waste",
    cat2 == "Waste emissions by gas (ktCO2e)",
    cat3 == "CH4") |>
  filter(between(year, ymd("2022-01-01"), ymd("2025-01-01"))) |>
  summarise(across(c(ktco2e), sum)) |>
  rename(emissions_ar4 = ktco2e) |>
  mutate(emissions_ar5 = emissions_ar4 / 25 * 28)

#Emissions in CPR with project still higher than Demonstration pathway in
#1st commitment period
```

The Demonstration pathway has lower waste methane emissions than the Current Policy Sceneraio with project for the first budget period - both using GWPs from IPCC AR4 as well as AR5.
